Augmentation of a Model Historical Harpsichord Mechanism for Haptic Enabled Interaction

### 2025-01-18 src/article.tex
Update article.tex
Updates from Overleaf
rm tightlist

\section{Abstract}\label{abstract}

@@

The project look at restoring tactile and haptic response of 17th
century harpsichords for an exhibition at the musical instrument museum
San Colombano in Bologna. San Colombano hosts the Tagliavini Collection
(from Luigi Ferdinando Tagliavini), a collection of approximately 70
historical keyboard instruments in working order
\textbackslash cite\{Tagliavini2009\}
(https://genusbononiae.it/en/san-colombano-tagliavini-collection/). Some
items in the collection are not in playing condition or are too fragile
to continue to be available for the public to play. One of those fragile
instruments in the Tagliavini collection is a 1547 Harpsichord by
Alessandro Trasuntino \textbackslash cite\{Treccani2019\}, which was
used as the basis for an interactive exhibition commisioned by San
Colombano.

@@

This is particularly pertinent as it relates to conservation of musical
instruments, as they represent both historic objects, but also tools for
creating music. Historical musical instruments present an interesting
problem for cultural heritage as they are both a physical object but
also a tool for making music. This begs the question, can we use haptics
to preserve experiences of the past? The proposed project intends to
develop further the methodologies implemented in musical haptics
projects {[}3{]} such as {[}2{]}. The final outcome will be to create a
framework for digital conservation of historic musical instruments such
as the 1638 Ioannes Ruckers and the 1646 Ioannes Couchet double-manual
harpsichords as referenced in the NEMUS project {[}4{]}.

@@

\begin{itemize}
\tightlist
\item
  existing sysetms

  \begin{itemize}
  \tightlist
  \item
    Yamaha Disklaviare
  \item
    Bosendorfer

    \begin{itemize}
    \tightlist
    \item
      290SE
    \item
      CEUS
    \end{itemize}
  \item
    Don Buchla / Moog Piano Bar
  \item
    PNOscan MIDI 9 QRS Music

    \begin{itemize}
    \tightlist
    \item
      PNOmation
    \end{itemize}
  \item
    Piano Disc
  \end{itemize}
\end{itemize}

@@

The paper by McAlpine \textbackslash cite\{McAlpine2014\} also has very
similar motivations to the project presented here. Both were started
from a musical instrument museum, housing a collection of historic
instruments gifted by a collector (Benton Fletcher / Taglivaini) who
wished for collection to remain playable, which is restricted growing
fragility of some items in the respective collections. Bothe projects
center around the creation of a custioom MIDI controller for interfacing
with sound synthesis engine. The keyboard consisted of two modified
commercial MIDI keyboard controllers in a two manual harpsichord
arrangement. Weherease the McAlpine keyboard used weighted keys, this
interface extends the idea by using the mechanism and strings under
tension to provide the haptic feedback. McAlpine posits a fully
paremterised haptic keyboard, but before taking such a leap we look at
what affect a more literal haptic keyboard may provide to the performer
experience.

The previous NIME project ``Tromba Moderna''
\textbackslash cite\{Baldwin2016\}, which looked at reconstructing a
full tromba marina to be used as a digital interface. The one of the
limitations flagged by McAlpine was the oneffectiveness of weighted keys
in mimicing the harpischord. The Tromba Moderna ``attempted to avoid
this problem by constructing a fully functioning replica of the tromba
marina'' \textbackslash cite\{Baldwin\}. This project applies the same
approach to the harpsichord.m where the projects deviate is a questions
of scale.

@@

Extension to \textbackslash cite\{McAlpine2014\} and
\textbackslash cite\{Baldwin2016\} are the open sourcing of hardware and
software for the project.

@@

\begin{itemize}
\tightlist
\item
  maintain problems
\item
  probe for research

  \begin{itemize}
  \tightlist
  \item
    on haptics how it affects performance
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  simple
\item
  aesthetics
\item
  split workload
\item
  parts

  \begin{itemize}
  \tightlist
  \item
    jack

    \begin{itemize}
    \tightlist
    \item
      quill
    \item
      jack body
    \item
      pivot
    \item
      tongue
    \end{itemize}
  \item
    key
  \item
    body
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  proof of concept
\item
  available on loan from museum
\item
  easy access to internal structure
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  for exhibition
\item
  problems of scale
\item
  problems of timescale
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  no electronics should be visible.
\item
  it should be robust and reliable, easily maintained by museum staff
  without complex technical processes
\item
  it should not augment or fix the mechanics, but present them with
  their original limitation
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Non-invasive: The sensor system should not require major modifications
  to the interface on which it is installed.
\item
  Low Latency: Reading sensors and then triggering an output should have
  a low latency. Using {[}paper on latency{]} a target was set as X ms
  with an upper limit set to 25ms beyond which the sensors would be
  deemed unusable
\item
  Reliable: Data obtained should be reliable and accurate enough not to
  result in false positives or false negatives without extensive
  filtering
\item
  Scalable: The system should be scalable both economically and
  temporally. Given the open soucre nature of the porject, it could not
  be prohibitvely expensive to carry out. Also any system that would
  work on a single key would have to scale financially to potentially 50
  or more. The time required to install and calibrate the sensors would
  also have to scale.
\item
  Expandable: Any sensor system should be functionally expandable to
  allow for the usage of the data for control over other parameters.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \textbackslash cite\{HapticKey\}
\item
  \textbackslash cite\{mcPherson1\} \textbackslash cite\{mcPherson2\}
\item
  \textbackslash cite\{Mudd2013\}
\item
  \textbackslash cite\{FritzClaudia2017Leon\}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Over arching themes

  \begin{itemize}
  \tightlist
  \item
    simplicity of design

    \begin{itemize}
    \tightlist
    \item
      Limited resources
    \item
      easily reproducible
    \end{itemize}
  \item
    availability and lead times
  \item
    iteration and design from distance
  \end{itemize}
\end{itemize}

@@

Following from the work in \textbackslash cite\{McPherson1\}
\textbackslash cite\{McPherson2\} we tested a system using the Fairchild
QRE1113 \textbackslash datasheetfigue. The QRE1113 is a combination
infrared LED and phototransistor sensitive to IR light. The
phototransistor of QRE1113 provides data on how much light is present
and in particular how much light is being relfected from a surface.
Since refelected light will be proportional to the distance of a surfcae
it is a good, close proximity, distance sensor. Distance is the way in
which the instruments in \textbackslash citeMcpherson1\&2 function as
well the Moog Piano Bar from which it too inspiration.

@@

\begin{itemize}
\tightlist
\item
  The triggering would need to be bespoke for every key as fluxuations
  in the dimensions of every piece mean that the threshold would be
  different.

  \begin{itemize}
  \tightlist
  \item
    Moving a problem into additional time in the calibtraryoin setup
    phase.
  \end{itemize}
\item
  Remebering the constraint that system isshould noyt fix problem, there
  would be no easy way to tell if a jack had nbot reseated.

  \begin{itemize}
  \tightlist
  \item
    In playing a harpihchord when the jack does not go to its rest
    position, not note would play. This fault should be carried over
    intp the final instrument.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  previous use
\item
  problem of what to sense
\item
  satisfied both criteria and design restrictions
\item
  usage

  \begin{itemize}
  \tightlist
  \item
    Sweet spot is around 6mm distance
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  linear gradient
\item
  on Jack
\item
  printed and cut on vinyl {[}find name{]}
\item
  cut to travel size
\item
  then to jack size
\item
  jacks varnished on side
\item
  alternatives

  \begin{itemize}
  \tightlist
  \item
    CNC plywood and mdf

    \begin{itemize}
    \tightlist
    \item
      problems
    \end{itemize}
  \item
    laser cut plywood and mdf

    \begin{itemize}
    \tightlist
    \item
      problems
    \end{itemize}
  \item
    3D printed jack body

    \begin{itemize}
    \tightlist
    \item
      problems
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Nano form factor small
\item
  allows for easy changes in chipset
\item
  ``Native USB''
  https://learn.adafruit.com/dude-where-s-my-com-port/native-usb-boards\#
\item
  BLE Model

  \begin{itemize}
  \tightlist
  \item
    Performative
  \item
    BLE capabilities for MIDI BLE transmission
  \item
    All analog pins available
  \item
    available at the time
  \item
    12-bit DAC
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  On board available, but volatile
\item
  retained across firmware changes
\item
  small,
\item
  could have been sd card or anything else
\item
  available
\item
  interfaced via rotary encoder
\end{itemize}

@@

The LEDs are on the same PCB as the sensors and are hidden when the
nameboard is in place.

@@

TLC5940 or other PWM driver such as in \textbackslash cite\{mcpherson\}
added complexity to assembling, but were avoid in favour of Addressable
LEDs with intergrated driver. The intergrated LEDs share the same data
line which has the limitation where a fualty LED will stop all following
LEDs from working. An individual PWM driver per PCB would allow avoid
situations where failutre cascades, but would add complexity in assembly
as the parts are typically SMD. Also, given the LEDs are fixed to the
board and could not be repaired in situ, the solution would be to
replace the PCB regardless of which approach was taken. Given the LEDs
were only used in debugging and calibration and otehrwise would not be
visible, a simpler design approach was taken.

\begin{itemize}
\tightlist
\item
  locating keys
\item
  displaying key state
\end{itemize}

@@

Transistor / Op-amp circuit could have been used similar to the one in
\textbackslash cite\{mcpherson\}.

@@

\begin{itemize}
\tightlist
\item
  limited ADC channels
\item
  ADC channels already multiplexed
\item
  simple
\item
  8 channel
\item
  THT
\item
  7 into 49

  \begin{itemize}
  \tightlist
  \item
    7 PCBS
  \item
    each with 7 sensors
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  EAGLE PCBs designed using EAGLE. Worth noting that AutoCAD has
  deprecated EAGLE and the programme will no longer be supported from
  2026 Project files will need to be transitioned to another standard,
  likely the KiCAD programme.
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    legs break easily
  \item
    end mount like A.P.Mc. time consuming
  \item
    problem

    \begin{itemize}
    \tightlist
    \item
      milling out keys
    \end{itemize}
  \end{itemize}
\item
  designed from distance
\item
  striped lines across pcbs
\item
  problems

  \begin{itemize}
  \tightlist
  \item
    additional terminals uneeded resulted in lots of additional work
  \item
    long cabling increase risk of breaking
  \end{itemize}
\item
  solutions

  \begin{itemize}
  \tightlist
  \item
    striping all adc lines with FFC
  \item
    solder pads to select channel
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  mount for nano to allow switching
\item
  initially push fit 2.54mm header

  \begin{itemize}
  \tightlist
  \item
    not a secure fit
  \end{itemize}
\item
  JST-PH used

  \begin{itemize}
  \tightlist
  \item
    crimping required custom cables

    \begin{itemize}
    \tightlist
    \item
      could be assembled from other cables
    \item
      length limited
    \end{itemize}
  \item
    prohibitevly expensive to crimpers
  \item
    3D print helper to allow for crimping with generic crimpers
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Push fit
  \item
    adjustable for best average placement
  \end{itemize}
\item
  Blinkers

  \begin{itemize}
  \tightlist
  \item
    reduce noise from other senors
  \end{itemize}
\item
  mounts

  \begin{itemize}
  \tightlist
  \item
    MacMini
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Arduino platform

  \begin{itemize}
  \tightlist
  \item
    library support and portability
  \item
    dependancy management
  \item
    wide spread use
  \end{itemize}
\item
  core algorithm

  \begin{itemize}
  \tightlist
  \item
    check components connected
  \item
    READ data from FRAM
  \item
    select key with rotary

    \begin{itemize}
    \tightlist
    \item
      click to edit, double to save
    \end{itemize}
  \item
    current threshold and reading sent over serial and plotted
  \item
    when over threshold send MIDI note
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Firmware
\item
  Keyboard CAD Data
\item
  PCB CAD Data
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  partial pre-assembly
\item
  handsolder components

  \begin{itemize}
  \tightlist
  \item
    analog lines
  \end{itemize}
\item
  connecting supports
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  in museum context
\item
  strings damped
\item
  headphones
\item
  interfacing with kontakt spitfire library
\item
  display

  \begin{itemize}
  \tightlist
  \item
    secure storage and access
  \item
    interface to the public
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  sensor surface

  \begin{itemize}
  \tightlist
  \item
    printed gradient
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  scale
\item
  time scale
\item
  working at distance
\item
  limitations

  \begin{itemize}
  \tightlist
  \item
    latency
  \item
    deprecation

    \begin{itemize}
    \tightlist
    \item
      software

      \begin{itemize}
      \tightlist
      \item
        EAGLE
      \end{itemize}
    \item
      hardware

      \begin{itemize}
      \tightlist
      \item
        FRAM
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Velocity, aftertouch
\item
  second jack row
\item
  reducing latency
\item
  alternative sensor surfaces
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  ESP-IDF

  \begin{itemize}
  \tightlist
  \item
    latency half
  \item
    workload high
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  avoid soldering cabling
\item
  power indicators on pcbs

  \begin{itemize}
  \tightlist
  \item
    debug if problem is power related Add power small power LEDs to the
    each PCB. It was not common for a board to not be functioning and
    thye result to be that the power weas not correctly connected.
    Especially when the ower is connected via somethgin such as an FFC
    cable where it is easy to plug-in the wrong way esepcially when
    fitting is to be carried out by someone unfmailiar wiyth the system.
    Power LEDs would quickly confirm whether the problem was power
    related. They would have to be small enough and low enough current
    draw to avoid glare or being mistaken with other indcator LEDs To
    continue using THT components, the choice would have to be moade
    carefully.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  {[}1538{]}
  https://carmentis.kmkg-mrah.be:443/eMP/eMuseumPlus?service=ExternalInterface\&module=collection\&objectId=106088\&viewType=detailView
\item
  {[}1531{]}https://museumcollections.rcm.ac.uk/collection/Details/collect/58
\end{itemize}

\section{Acknowledgments}\label{acknowledgments}

A project of this breadth developed across 3 institutes and 2 countries
cannot be carried without the support and technical knowledge of many
skilled and passionate people.

We acknowledge the invaluable contributions of

\begin{itemize}
\tightlist
\item
  Unibo

  \begin{itemize}
  \tightlist
  \item
    Alma Labor

    \begin{itemize}
    \tightlist
    \item
      Massimiliano Fraulini and the staff of Alma Labor at the Yiversità
      of bologna for their 3D printing expertise and generosity with
      electronics equipment
    \end{itemize}
  \item
    Fisica Tecnica

    \begin{itemize}
    \tightlist
    \item
      Maurzio Chendi Fabrizio Casarini, Giulio Severo and Barbara
      Costantini
    \end{itemize}
  \end{itemize}
\item
  San Colombano

  \begin{itemize}
  \tightlist
  \item
    Catalina Vincens
  \item
    Roberto Livi
  \end{itemize}
\item
  Edinburgh

  \begin{itemize}
  \tightlist
  \item
    uCreate

    \begin{itemize}
    \tightlist
    \item
      Simeon, Stuart, Millie, and Ruth of uCreate at the Univeristy of
      Edinburgh for their support with soldering facilities.
    \end{itemize}
  \item
    ECA

    \begin{itemize}
    \tightlist
    \item
      Richard Collins and the Digital Making Staff at the Edinburgh
      College of Art for 3D printing instruction and CNC support.
    \item
      Joe Hathaway of Digital Development at the College of Art for
      providing electronics components
    \item
      St Cecilia's Jenny Nex
    \end{itemize}
  \end{itemize}
\end{itemize}

>>>>

@@

The project look at restoring tactile and haptic response of 17th
century harpsichords for an exhibition at the musical instrument museum
San Colombano in Bologna. San Colombano hosts the Tagliavini Collection
(from Luigi Ferdinando Tagliavini), a collection of approximately 70
historical keyboard instruments in working order \cite{Tagliavini2007}
(https://genusbononiae.it/en/san-colombano-tagliavini-collection/). Some
items in the collection are not in playing condition or are too fragile
to continue to be available for the public to play. One of those fragile
instruments in the Tagliavini collection is a 1547 Harpsichord by
Alessandro Trasuntino \cite{Wraight2024}, which was used as the basis
for an interactive exhibition commisioned by San Colombano.

@@

This is particularly pertinent as it relates to conservation of musical
instruments, as they represent both historic objects, but also tools for
creating music. Historical musical instruments present an interesting
problem for cultural heritage as they are both a physical object but
also a tool for making music. This begs the question, can we use haptics
to preserve experiences of the past? The proposed project intends to
develop further the methodologies implemented in musical haptics
projects \cite{MusicalHaptics} such as \cite{Timmermans2020}.

@@

\begin{itemize}
\item
  existing sysetms

  \begin{itemize}
  \item
    Yamaha Disklavier used as a basis for exploration into haptics and
    the piano keyabord
    \cite{MusicalHaptics2018_04, MusicalHaptics2018_05,
    MusicalHaptics2018_13}
  \item
    Bosendorfer: LED photo transistor pair \cite{Moog1990} for obtaining
    performance data \cite{MusicalHaptics2018_05}

    \begin{itemize}
    \item
      290SE
    \item
      CEUS
    \end{itemize}
  \item
    Don Buchla / Moog Piano Bar
  \item
    PNOscan MIDI 9 QRS Music as MIDI Piano system \cite{McPherson2013}

    \begin{itemize}
    \item
      PNOmation
    \end{itemize}
  \item
    Piano Disc
  \end{itemize}
\end{itemize}

@@

The paper by McAlpine \cite{McAlpine2014} also has very similar
motivations to the project presented here. Both were started from a
musical instrument museum, housing a collection of historic instruments
gifted by a collector (Benton Fletcher / Taglivaini) who wished for
collection to remain playable, which is restricted growing fragility of
some items in the respective collections. Bothe projects center around
the creation of a custioom MIDI controller for interfacing with sound
synthesis engine. The keyboard consisted of two modified commercial MIDI
keyboard controllers in a two manual harpsichord arrangement. Weherease
the McAlpine keyboard used weighted keys, this interface extends the
idea by using the mechanism and strings under tension to provide the
haptic feedback. McAlpine posits a fully paremterised haptic keyboard,
but before taking such a leap we look at what affect a more literal
haptic keyboard may provide to the performer experience.

The previous NIME project ``Tromba Moderna'' \cite{Baldwin2016}, which
looked at reconstructing a full tromba marina to be used as a digital
interface. The one of the limitations flagged by McAlpine was the
oneffectiveness of weighted keys in mimicing the harpischord. The Tromba
Moderna ``attempted to avoid this problem by constructing a fully
functioning replica of the tromba marina'' \cite{Baldwin2016}. This
project applies the same approach to the harpsichord.m where the
projects deviate is a questions of scale.

@@

Extension to \cite{McAlpine2014} and \cite{Baldwin2016} are the open
sourcing of hardware and software for the project.

@@

\begin{itemize}
\item
  maintain problems
\item
  probe for research

  \begin{itemize}
  \item
    on haptics how it affects performance
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  simple
\item
  aesthetics
\item
  split workload
\item
  parts of harpsichord

  \begin{itemize}
  \item
    jack

    \begin{itemize}
    \item
      quill
    \item
      jack body
    \item
      pivot
    \item
      tongue
    \end{itemize}
  \item
    key
  \item
    body
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  proof of concept
\item
  available on loan from museum
\item
  easy access to internal structure
\end{itemize}

@@

\begin{itemize}
\item
  for exhibition
\item
  problems of scale
\item
  problems of timescale
\end{itemize}

@@

\begin{itemize}
\item
  no electronics should be visible.
\item
  it should be robust and reliable, easily maintained by museum staff
  without complex technical processes
\item
  it should not augment or fix the mechanics, but present them with
  their original limitation
\end{itemize}

@@

\begin{itemize}
\item
  Non-invasive: The sensor system should not require major modifications
  to the interface on which it is installed.
\item
  Low Latency: Reading sensors and then triggering an output should have
  a low latency. Using {[}paper on latency{]} a target was set as X ms
  with an upper limit set to 25ms beyond which the sensors would be
  deemed unusable
\item
  Reliable: Data obtained should be reliable and accurate enough not to
  result in false positives or false negatives without extensive
  filtering
\item
  Scalable: The system should be scalable both economically and
  temporally. Given the open soucre nature of the porject, it could not
  be prohibitvely expensive to carry out. Also any system that would
  work on a single key would have to scale financially to potentially 50
  or more. The time required to install and calibrate the sensors would
  also have to scale.
\item
  Expandable: Any sensor system should be functionally expandable to
  allow for the usage of the data for control over other parameters.
\end{itemize}

@@

\begin{itemize}
\item
  \cite{Timmermans2020}
\item
  \cite{McPherson2013} \cite{McPherson2019}
\item
  \cite{Mudd2013}
\item
  \cite{Fritz2017}
\end{itemize}

@@

\begin{itemize}
\item
  Over arching themes

  \begin{itemize}
  \item
    simplicity of design

    \begin{itemize}
    \item
      Limited resources
    \item
      easily reproducible
    \end{itemize}
  \item
    availability and lead times
  \item
    iteration and design from distance
  \end{itemize}
\end{itemize}

@@

Following from the work in \cite{McPherson2013, McPherson2019} we tested
a system using the Fairchild QRE1113 \textbackslash datasheetfigue. The
QRE1113 is a combination infrared LED and phototransistor sensitive to
IR light. The phototransistor of QRE1113 provides data on how much light
is present and in particular how much light is being relfected from a
surface. Since refelected light will be proportional to the distance of
a surfcae it is a good, close proximity, distance sensor. Distance is
the way in which the instruments in \cite{McPherson2013, McPherson2019}
function as well the Moog Piano Bar from which it too inspiration.

@@

\begin{itemize}
\item
  The triggering would need to be bespoke for every key as fluxuations
  in the dimensions of every piece mean that the threshold would be
  different.

  \begin{itemize}
  \item
    Moving a problem into additional time in the calibtraryoin setup
    phase.
  \end{itemize}
\item
  Remebering the constraint that system isshould noyt fix problem, there
  would be no easy way to tell if a jack had nbot reseated.

  \begin{itemize}
  \item
    In playing a harpihchord when the jack does not go to its rest
    position, not note would play. This fault should be carried over
    intp the final instrument.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  previous use
\item
  problem of what to sense
\item
  satisfied both criteria and design restrictions
\item
  usage

  \begin{itemize}
  \item
    Sweet spot is around 6mm distance
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  linear gradient
\item
  on Jack
\item
  printed and cut on vinyl {[}find name{]}
\item
  cut to travel size
\item
  then to jack size
\item
  jacks varnished on side
\item
  alternatives

  \begin{itemize}
  \item
    CNC plywood and mdf

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \item
    laser cut plywood and mdf

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \item
    3D printed jack body

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Nano form factor small
\item
  allows for easy changes in chipset
\item
  ``Native USB''
  https://learn.adafruit.com/dude-where-s-my-com-port/native-usb-boards\#
\item
  BLE Model

  \begin{itemize}
  \item
    Performative
  \item
    BLE capabilities for MIDI BLE transmission
  \item
    All analog pins available
  \item
    available at the time
  \item
    12-bit DAC
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  On board available, but volatile
\item
  retained across firmware changes
\item
  small,
\item
  could have been sd card or anything else
\item
  available
\item
  interfaced via rotary encoder
\end{itemize}

@@

The LEDs are on the same PCB as the sensors and are hidden when the
nameboard is in place.2024

@@

TLC5940 or other PWM driver such as in \cite{McPherson2013} added
complexity to assembling, but were avoid in favour of Addressable LEDs
with intergrated driver. The intergrated LEDs share the same data line
which has the limitation where a fualty LED will stop all following LEDs
from working. An individual PWM driver per PCB would allow avoid
situations where failutre cascades, but would add complexity in assembly
as the parts are typically SMD. Also, given the LEDs are fixed to the
board and could not be repaired in situ, the solution would be to
replace the PCB regardless of which approach was taken. Given the LEDs
were only used in debugging and calibration and otehrwise would not be
visible, a simpler design approach was taken.

\begin{itemize}
\item
  locating keys
\item
  displaying key state
\end{itemize}

@@

Transistor / Op-amp circuit could have been used similar to the one in
\cite{McPherson2013}.

@@

\begin{itemize}
\item
  limited ADC channels
\item
  ADC channels already multiplexed
\item
  simple
\item
  8 channel
\item
  THT
\item
  7 into 49

  \begin{itemize}
  \item
    7 PCBS
  \item
    each with 7 sensors
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  EAGLE PCBs designed using EAGLE. Worth noting that AutoCAD has
  deprecated EAGLE and the programme will no longer be supported from
  2026 Project files will need to be transitioned to another standard,
  likely the KiCAD programme.
\end{itemize}

@@

  \begin{itemize}
  \item
    legs break easily
  \item
    end mount like A.P.Mc. time consuming
  \item
    problem

    \begin{itemize}
    \item
      milling out keys
    \end{itemize}
  \end{itemize}
\item
  designed from distance
\item
  striped lines across pcbs
\item
  problems

  \begin{itemize}
  \item
    additional terminals uneeded resulted in lots of additional work
  \item
    long cabling increase risk of breaking
  \end{itemize}
\item
  solutions

  \begin{itemize}
  \item
    striping all adc lines with FFC
  \item
    solder pads to select channel
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  mount for nano to allow switching
\item
  initially push fit 2.54mm header

  \begin{itemize}
  \item
    not a secure fit
  \end{itemize}
\item
  JST-PH used

  \begin{itemize}
  \item
    crimping required custom cables

    \begin{itemize}
    \item
      could be assembled from other cables
    \item
      length limited
    \end{itemize}
  \item
    prohibitevly expensive to crimpers
  \item
    3D print helper to allow for crimping with generic crimpers
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \item
    Push fit
  \item
    adjustable for best average placement
  \end{itemize}
\item
  Blinkers

  \begin{itemize}
  \item
    reduce noise from other senors
  \end{itemize}
\item
  mounts

  \begin{itemize}
  \item
    MacMini
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Arduino platform

  \begin{itemize}
  \item
    library support and portability
  \item
    dependancy management
  \item
    wide spread use
  \end{itemize}
\item
  core algorithm

  \begin{itemize}
  \item
    check components connected
  \item
    READ data from FRAM
  \item
    select key with rotary

    \begin{itemize}
    \item
      click to edit, double to save
    \end{itemize}
  \item
    current threshold and reading sent over serial and plotted
  \item
    when over threshold send MIDI note
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Firmware
\item
  Keyboard CAD Data
\item
  PCB CAD Data
\end{itemize}

@@

\begin{itemize}
\item
  partial pre-assembly
\item
  handsolder components

  \begin{itemize}
  \item
    analog lines
  \end{itemize}
\item
  connecting supports
\end{itemize}

@@

\begin{itemize}
\item
  in museum context
\item
  strings damped
\item
  headphones
\item
  interfacing with kontakt spitfire library
\item
  display

  \begin{itemize}
  \item
    secure storage and access
  \item
    interface to the public
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  sensor surface

  \begin{itemize}
  \item
    printed gradient
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  scale
\item
  time scale
\item
  working at distance
\item
  limitations

  \begin{itemize}
  \item
    latency
  \item
    deprecation

    \begin{itemize}
    \item
      software

      \begin{itemize}
      \item
        EAGLE
      \end{itemize}
    \item
      hardware

      \begin{itemize}
      \item
        FRAM
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Velocity, aftertouch
\item
  second jack row
\item
  reducing latency
\item
  alternative sensor surfaces
\end{itemize}

@@

\begin{itemize}
\item
  ESP-IDF

  \begin{itemize}
  \item
    latency half
  \item
    workload high
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  avoid soldering cabling
\item
  power indicators on pcbs

  \begin{itemize}
  \item
    debug if problem is power related Add power small power LEDs to the
    each PCB. It was not common for a board to not be functioning and
    thye result to be that the power weas not correctly connected.
    Especially when the ower is connected via somethgin such as an FFC
    cable where it is easy to plug-in the wrong way esepcially when
    fitting is to be carried out by someone unfmailiar wiyth the system.
    Power LEDs would quickly confirm whether the problem was power
    related. They would have to be small enough and low enough current
    draw to avoid glare or being mistaken with other indcator LEDs To
    continue using THT components, the choice would have to be moade
    carefully.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  {[}1538{]}
  https://carmentis.kmkg-mrah.be:443/eMP/eMuseumPlus?service=ExternalInterface\&module=collection\&objectId=106088\&viewType=detailView
\item
  {[}1531{]}https://museumcollections.rcm.ac.uk/collection/Details/collect/58
\end{itemize}
